Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-04-06 Manuscript/manuscript.Rmd
manuscript edits

We examined the class imbalance of the different data resolutions for each cell line. As expected, the total number of genomic bins decreased as resolution decreased. That is, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins (Figure 4 (A)). The imbalance ratios, however, remained relatively stable (Figure 4 (B)). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Ideally, a balanced dataset will have an IR close to 0.50. Here, the low imbalance ratios indicated a wide disparity between majority and minority classes (i.e. a genomic bin was much more likely to not overlap with a flanked TAD boundary than it was to overlap with one). Therefore suitable resampling techniques were needed.

@@

## CLASS IMBALANCE SEVERELY IMPACTS PERFORMANCE OF MODELS WITH OVERLAP COUNTS PREDICTOR TYPES

We first evaluated model performances using only the overlap counts predictor types without re-sampling as these characteristics represented models with the least granularity and could act as a baseline. We measured the performances across resolutions, for each cell line, and found that .

Some notable results became clear. First, we see that the conventional AUC value was unable to accurately portray the poor performance of the models when compared to the composite metrics F1-score, MCC, and, to a lesser extent, the balanced accuracy. Inspecting the F1-score and MCC more closely, we see that some models at lower resolutions (for either cell line) were unable to correctly predict any bins that overlap with flanked TAD boundaries in the test data, as indicated by the missing bars.

## DATA RE-SAMPLING IMPROVES MODEL PERFORMANCE

Comparisons of the effects of the various resampling techniques were then made across resolutions, for each cell line. Results showed that performances drastically improved among the composite metrics (Figure 8 (B-D)). Models implemented on randomly under-sampled and SMOTE-sampled data performed similarly, while performances on randomly over-sampled data lagged behind, yet still outperformed models with no data re-sampling. Overall, there appeared the be a downward trend in model performance as resolutions decreased, across both cell lines.

>>>>

We examined the class imbalance of the different data resolutions for each cell line. As expected, the total number of genomic bins decreased as resolution decreased. That is, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins (Figure 4A). The imbalance ratios, however, remained relatively stable (Figure 4B). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Ideally, a balanced dataset will have an IR close to 0.50. Here, the low imbalance ratios indicated a wide disparity between majority and minority classes (i.e. a genomic bin was much more likely to not overlap with a flanked TAD boundary than it was to overlap with one). Therefore suitable resampling techniques were needed.

@@

## CLASS IMBALANCE SEVERELY IMPACTS PERFORMANCE AMONG MODELS WITH EITHER OVERLAP COUNTS OR PERCENTS PREDICTOR TYPES

We first evaluated model performances for the GM12878 cell line, using only the overlap counts predictor types, without re-sampling, as these characteristics represented models with the least granularity and could act as a baseline. We measured the performances across resolutions, and found that a conventional measure such as AUC artificially inflated model performances (Figure 5A (I)). In evaluating the composite metrics like F1-Score and MCC, we found more reasonable depictions of poor performance, given such high levels of class imbalance (Figure 5A (II-IV)). At the 50 kb and 100 kb resolution, for overlap counts, the F1-Score and MCC could not be determined given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data. When including overlap percents perdictor types in the models, we found that AUC values were comparable to OC models, across resolution. However, when considering both the F1-Score and MCC, we found that OP models performed worse than OC models. Similarly, at the 50 kb and 100 kb resolutions, models built on OP were unable produce F1-Scores or MCC values. Similar results were found for the K562 cell line (Supplementary Figure 10A (I-IV)).

## DATA RE-SAMPLING IMPROVES PERFORMANCE AMONG MODELS WITH OVERLAP COUNTS PREDICTOR TYPES

Comparisons of the effects of the various resampling techniques were then made for models using overlap counts predictor types, across resolutions. Simply considering AUC did not depict any notable improvement in performances, regardless of the re-sampling technique (Figure 5B-5D (I)). However, when considering the other composite metrics we found that there was a substantial increase in F1-Score and MCC in each of the 3 resampling techniques. For overlap counts predictors, both RUS and SMOTE resampled datasets were found to yield consistently better results than no resampling and ROS.

Results showed that performances drastically improved among the composite metrics (Figure 8 (B-D)). Models implemented on randomly under-sampled and SMOTE-sampled data performed similarly, while performances on randomly over-sampled data lagged behind, yet still outperformed models with no data re-sampling. Overall, there appeared the be a downward trend in model performance as resolutions decreased, across both cell lines.
